---
title: "Downstream Analysis Notebook"
author: "Dídac Jiménez"
date: "2022-11-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preparation and setup

### Libraries

Load libraries used in this notebook., indicate the preprocessing tool used, the species, and the path to the files.

```{r echo=FALSE, warning=FALSE}
library(Seurat, quietly = TRUE)
library(tidyverse, quietly = TRUE)
library(Matrix, quietly = TRUE) # load libraries
library(DropletUtils, quietly = TRUE)
```

### Set parameters

Indicate the preprocessing tool used, the species, and the path to the files.

```{r}
projectname<-"Prototype"
tool<-"KallistoBustools" # "KallistoBustools", "UmiTools or "SalmonAlevin
matrixpath<-"~/Work/Runs/KB_R5/output/counts_unfiltered/cells_x_genes.mtx"
featurespath<-"~/Work/Runs/KB_R5/output/counts_unfiltered/cells_x_genes.genes.txt" #Only for KB
cellspath<-"~/Work/Runs/KB_R5/output/counts_unfiltered/cells_x_genes.barcodes.txt" #Only for KB
MT_genespath<-"~/Work/Refs/Human/MTgenes"
```

### Load raw files into counts matrix

Outputs from each preprocessing tool are different. I will define and use different functions to generate a counts matrix.

```{r}
if (tool == "KallistoBustools"){
  counts<-ReadMtx(
    mtx = matrixpath,
    cells = cellspath,
    features = featurespath,
    cell.column = 1,
    feature.column = 1,
    cell.sep = "\t",
    feature.sep = "\t",
    skip.cell = 0,
    skip.feature = 0,
    mtx.transpose = TRUE,
    unique.features = TRUE,
    strip.suffix = FALSE
)
  # Ensembl version removal
  rownames(counts)<-sub('\\.[0-9]*$', '', rownames(counts))
}

if (tool == "SalmonAlevin"){
  txi <- tximport::tximport(matrixpath, type="alevin")
  counts<-txi$counts
  # Ensembl version removal
  rownames(counts)<- sub('\\.[0-9]*$', '', rownames(txi$counts))
}

if (tool == "UmiTools"){
  counts <- Read10X(sub("matrix.mtx.gz","",matrixpath), 
                  gene.column = 1)
}

```

### Tag mitochondrial genes

Mitochondrial genes must have a tag on their name so that we can use the proportion of mitochondrial transcripts as a quality control measure. This is better done before the construction of the `Seurat` object. For this we will use a list of all mitochondrial genes (species dependent!) previously acquired. 

```{r}
#List of the names of our genes
gene_names<-data.frame(ens=rownames(counts))

# List generated Using Biomart from Ensembl to  create a list of ENS genes corresponding to mitochondrial genes 
MT_genes<-data.frame(MT=read_csv(MT_genespath, show_col_types = FALSE))

# Loop through all gene names and tag those found in the list
MT_count <-0
for (i in 1:length(gene_names$ens)){
  if (gene_names$ens[i] %in% MT_genes$ensembl_gene_id){
    MT_count<-MT_count+1 # To check if it works
    gene_names$ens[i] <- paste(c("MT-",gene_names$ens[i]),sep="",collapse = "")
  }
}

# Change row names of counts matrix 
row.names(counts)<-gene_names$ens

# Remove non useful objects
#rm(gene_names,MT_genes,MT_count)
```


### Generate Seurat object

Finally we create the 'Seurat' object which we will use to proceed with the analysis.

```{r}
seurat.obj <- CreateSeuratObject(counts = counts, 
                                 project = "projectname", 
                                 min.cells = 1, 
                                 min.features = 1)
```

## Quality Control

### Data exploration

First we compute the percentage of mitochondrial transcripts for each cell, to use in later filtering. 

```{r}
# We use the label we added earlier and add the percentage as a column in the metadata of each cell
seurat.obj[["percent.mt"]]<-PercentageFeatureSet(seurat.obj,pattern = "MT")
```

We then plot the three metadata values: Mitochondrial percentage, number of molecules and number of features. A good dataset should have linear relationship between number of features and number of molecules, though a plateau is fine. As for the violin plot, we will use it to set the limits for our filtering.

```{r}
VlnPlot(seurat.obj, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3)
FeatureScatter(seurat.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")+geom_smooth(method='lm')
```

### Filtering

We filter the cells by setting a limit on:
*Mitochondrial percentage: An upper limit, high values may indicate a broken nucleus, thus a compromised cell
*Number of molecules: An upper limit, very high values may indicate the droplet contained two cells
*Number of features: A lower limit, droplets with too few features may only contein environmental RNA

These filters depend on each dataset and should be changed on each execution.

```{r}
seurat.obj <- subset(seurat.obj,
                     subset=nFeature_RNA >500 &nCount_RNA < 3000 & percent.mt <10)
```

We make the violin plots again to check the filtering:

```{r}
VlnPlot(seurat.obj, features = c("nCount_RNA","nFeature_RNA","percent.mt"), ncol = 3)
FeatureScatter(seurat.obj, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")+geom_smooth(method='lm')
```

## Notmalization and scaling

### Normalization

Normalize the values so we can make comparisons across cells. Each value is divided by the total expression in its cell, multiplied by a scale
factor and then log transformed.

```{r}
seurat.obj<-NormalizeData(seurat.obj,normalization.method = "LogNormalize", scale.factor = 10000)
```

### Highly variable features

We do this to focus our analysis on the genes that differentiate our cells.

```{r}
seurat.obj <- FindVariableFeatures(seurat.obj, selection.method = "vst", nfeatures = 2000)
```

### Scale data

We run `ScaleData()` to shift the expression of each gene, so that the mean expression across cells is 0 and scales the expression of each gene, so that the variance across cells is 1. This step gives equal weight in downstream analyses, so that highly-expressed genes do not dominate and is a required step prior to dimensional reduction techniques.

```{r}
gene_names <- rownames(seurat.obj)
seurat.obj <- ScaleData(seurat.obj, features = gene_names)
```

## Dimension reduction

### Principal Component Analysis

We run the analysis with `RunPCA()`. 

```{r}
seurat.obj <- RunPCA(seurat.obj, features =VariableFeatures(seurat.obj))
```

We make an elbow plot. We want to keep enough PC's to capture most of the variation

```{r}
ElbowPlot(seurat.obj)
```

### Clustering

Here we specify how many PCA's to use and the resolutions for the cluster.

```{r}
seurat.obj <- FindNeighbors(seurat.obj,dims = 1:10)
seurat.obj <- FindClusters(seurat.obj,resolution = c(0.05,0.4,0.6,0.8,1))
DimPlot(seurat.obj, group.by = "RNA_snn_res.0.8", label = TRUE)
```























